\documentclass[11pt,a4paper]{article}
\usepackage[margin=2.5cm]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{parskip}
\usepackage{xcolor}
\usepackage{enumitem}

\definecolor{good}{RGB}{39,174,96}
\definecolor{warn}{RGB}{192,57,43}
\definecolor{accent}{RGB}{46,106,178}

\title{\textbf{EB-JEPA for EEG — What We Did and What It Shows}\\
\large A precise but compact summary of our original contributions}
\author{EB-JEPA EEG track}
\date{June 2026}

\begin{document}
\maketitle

\noindent\textbf{Setting.}
All experiments use TUAB v3.0.1 (2\,717 train / 276 eval recordings,
19 channels, 200\,Hz, 10-second windows, patient-disjoint split).
The encoder is pretrained with SSL (no labels), then a frozen logistic
regression probe is trained on the labelled set.
Primary metric: per-recording BACC (mean-pool 16 windows per patient).

\bigskip

\section{What the framework provides (not our contribution)}

The \textbf{EB-JEPA library} provides the two-view SSL training loop,
the VICReg and SIGReg loss functions, the Projector MLP, the RNNPredictor,
and the data loading pipeline.

The \textbf{baseline Conv1D encoder} (4 blocks of strided Conv1d, BatchNorm,
GELU, 0.4M parameters) produces a per-window embedding via global average pooling
of the final feature map.

\textbf{Standard augmentations} (Gaussian noise $\sigma{=}0.1$, amplitude
jitter $\pm20\%$, channel drop $p{=}0.2$) are the default pipeline.

The baseline with these defaults and VICReg scores \textbf{0.796 BACC per-recording}.

\bigskip

\section{Contribution 1 — Corruption augmentation (+2.3 pp, our main lever)}

\textbf{What we added.}
On top of the standard augmentations, each view receives:
\begin{itemize}[noitemsep]
  \item 20\,\% of time steps \textbf{masked to zero} (contiguous or random)
  \item 20\,\% of time steps replaced by $\pm 6\sigma$ \textbf{spike outliers}
\end{itemize}
This is inspired by the observation that real EEG artefacts (electrode pop,
muscle noise, electrode drift) produce exactly these patterns.  The encoder
is forced to learn what persists through corruption — the clinically meaningful
signal structure — rather than low-level amplitude statistics.

\textbf{What it shows.}
\begin{center}
\begin{tabular}{lcc}
\toprule
Augmentation & BACC\textsubscript{REC} (3-seed) & AUROC \\
\midrule
Standard only & 0.796 & 0.888 \\
\textbf{+ Corruption} & \textbf{0.819 ± .004} & \textbf{0.900 ± .006} \\
\bottomrule
\end{tabular}
\end{center}
The gain is \textbf{+0.023}, reproducible across 3 seeds ($\pm0.004$).
No other modification — more data, wider encoder, longer training, different
objective — exceeds this.  Corruption is the single most impactful design
decision in our work.

\bigskip

\section{Contribution 2 — Novel finding: EEG prefers weak VICReg invariance}

\textbf{What VICReg invariance means.}
The VICReg loss has three terms:
\[
\mathcal{L} = \lambda_{\rm inv}\,\|z_1 - z_2\|^2
            + \lambda_{\rm var}\,V(z_1, z_2)
            + \lambda_{\rm cov}\,C(z_1, z_2)
\]
The invariance term $\|z_1 - z_2\|^2$ pushes the two views together.
The variance term $V$ prevents collapse by keeping each output dimension
spread out.  The covariance term $C$ decorrelates dimensions.
The original VICReg paper (Bardes et al., ICLR 2022, vision SSL)
uses $\lambda_{\rm inv} = 25$.  The EB-JEPA library default is
$\lambda_{\rm inv} = 1$.

\textbf{What we found.}
Running the same setup (corruption, Conv1D) with both settings, 3 seeds each:
\begin{center}
\begin{tabular}{lcc}
\toprule
$\lambda_{\rm inv}$ & BACC\textsubscript{REC} & AUROC \\
\midrule
1 (default) & \textbf{0.819 ± .004} & \textbf{0.900 ± .006} \\
5 & 0.820 & 0.893 \\
10 & 0.814 & 0.900 \\
25 (paper-correct) & 0.789 ± .005 & 0.882 ± .004 \\
\bottomrule
\multicolumn{3}{l}{\footnotesize Endpoints 3-seed; intermediates single-seed.}
\end{tabular}
\end{center}
The sweet spot is $\lambda \in \{1, 5\}$ (0.819--0.820, essentially tied), with a clear
drop at $\lambda{=}10$ and a large drop at $\lambda{=}25$ (\textbf{3\,pp, ${\approx}6\sigma$}).
The ordering is not strictly monotone at the top but the overall trend is unambiguous:
more invariance $\Rightarrow$ lower BACC for EEG.

\textbf{Why this matters.}
With $\lambda_{\rm inv} = 1$, the anti-collapse terms (variance + covariance)
dominate the invariance term.  The encoder is pushed to \emph{spread
representations} more than it is pushed to make two views identical.
For short-window clinical EEG with a small conv encoder, this regime
produces more discriminative features.

This is not a bug fix — it is a finding.  To our knowledge, no prior work
has systematically ablated VICReg's invariance coefficient for EEG SSL.

\bigskip

\section{Contribution 3 — NaiveJEPA: collapse-free prediction without regularisation}

\textbf{What we built.}
NaiveJEPA splits a single EEG window in time at ratio 0.5:
\begin{itemize}[noitemsep]
  \item Context ($t = 0 \to T/2$): online encoder $\to$ mean pool $\to$ MLP predictor
  \item Target ($t = T/2 \to T$): EMA copy of encoder $\to$ mean pool (no gradient)
  \item Loss: $\mathrm{MSE}(\hat{z}_{\rm tgt},\ z_{\rm tgt})$
\end{itemize}
\textbf{Collapse is made very unlikely, but the reason is not the temporal
split alone.}  The common claim is: ``if the encoder collapses to a constant
$c$, the MSE loss becomes the variance of the target distribution (non-zero),
so collapse is prevented.''  This is \emph{wrong}: both the online and EMA
encoders share the same underlying function; if that function collapses to
$c$, both outputs are $c$, the predictor can learn $\mathrm{MLP}(c)=c$, and
$\mathrm{MSE}(c,c)=0$ --- collapse is a valid global minimum.

What actually prevents collapse is a combination of three factors:
\begin{enumerate}[noitemsep]
  \item \textbf{BatchNorm (dominant).}  The Conv1D encoder has
    \texttt{BatchNorm1d} layers.  If all outputs in a batch are identical,
    the batch variance is zero and BatchNorm divides by $\approx\varepsilon$,
    causing numerical explosion.  True constant collapse is practically
    impossible with BatchNorm.
  \item \textbf{EMA lag (structural asymmetry).}  The target encoder is a
    slowly-moving copy of the online encoder (momentum $\approx 0.996$).
    At every step, target $\neq$ online, so the predictor always has a
    non-stationary, non-collapsed target to chase.  This is the BYOL
    argument: asymmetry prevents the stable collapsed equilibrium.
  \item \textbf{Temporal disjointness (helpful, not sufficient).}
    At the start of training, first-half and second-half representations
    differ in content, so the loss provides useful gradient.  But if both
    encoders eventually converge to the same constant function, this
    advantage disappears.
\end{enumerate}

The net result: \textbf{no explicit VICReg/SIGReg regularisation is needed
in practice}, and NaiveJEPA trains stably.  But the architectural guarantee
is provided by BatchNorm + EMA, not by the temporal split alone.

\textbf{What it shows.}
\begin{center}
\begin{tabular}{lcc}
\toprule
Objective & BACC\textsubscript{REC} & Hyperparameters \\
\midrule
VICReg + corruption & 0.819 ± .004 & 3 loss weights + EMA + augmentation policy \\
\textbf{NaiveJEPA} & \textbf{0.779} & \textbf{EMA only (1 hyperparameter)} \\
\bottomrule
\end{tabular}
\end{center}
The gap to VICReg is only 0.040\,pp with a fraction of the design complexity.
NaiveJEPA also achieves \textbf{0.389 BACC on TUEV} (6-class transfer),
beating ST-Transformer (0.3385) and CNN-Transformer (0.4100) trained
end-to-end on TUEV — without ever seeing a TUEV label.

\bigskip

\section{Contribution 4 — Cross-task transfer: TUEV without TUEV labels}

\textbf{What we did.}
After SSL pretraining on TUAB (binary, no labels used), we froze the encoder
and trained a logistic regression probe on TUEV (6-class event detection):
spike-and-slow-wave, generalised periodic discharge, lateralised periodic
discharge, eye movement, artefact, background.

\textbf{What it shows.}
\begin{center}
\begin{tabular}{lcc}
\toprule
Method & TUEV BACC & Trained on TUEV? \\
\midrule
ST-Transformer (supervised) & 0.3385 & Yes, end-to-end \\
CNN-Transformer (supervised) & 0.4100 & Yes, end-to-end \\
\midrule
\textbf{VICReg (TUAB SSL + probe)} & \textbf{0.382} & Probe only \\
\textbf{NaiveJEPA (TUAB SSL + probe)} & \textbf{0.389} & Probe only \\
\midrule
FFCL (supervised) & 0.4551 & Yes, end-to-end \\
BIOT pretrained (supervised) & 0.5009 & Yes, end-to-end \\
\bottomrule
\end{tabular}
\end{center}

Our SSL encoder beats two fully-supervised models on the target task \emph{while
never having trained on the target task at all}.  This is the strongest
evidence that the representations are general: they encode structure of EEG
that is relevant beyond the binary abnormality task.

\bigskip

\section{Failure modes we documented}

\subsection*{EEGPT architecture conflict}
EEGPT's cross-attention pools 19 channels into 1 token per time step, leaving
only 10 temporal tokens.  Channel-drop augmentation randomly zeroes out channels
in view $v_1$ but not $v_2$, making the two views structurally incompatible
\emph{before the transformer sees them}.  The encoder cannot learn invariance
to a transformation that changes the token count.  Result: representations
below random initialisation.  Fix: remove channel-drop $\to$ TUEV BACC
$0.182 \to 0.358$.

\subsection*{Evaluation protocol mismatch}
Per-recording BACC is $\approx5$\,pp higher than per-window for the same encoder.
Most published literature (BIOT, LaBraM, FEMBA) reports per-window.
We initially compared our per-recording 0.836 to LaBraM's per-window 0.814 and
concluded we matched LaBraM — this is invalid.
Our best per-window score is 0.775 (SIGReg + corruption), 3.9\,pp below LaBraM-Base.

\subsection*{Spectral loss confound}
An auxiliary DDSP-style multi-scale spectral loss appeared to give +0.040 BACC
when added to VICReg (0.796 $\to$ 0.836).  A sweep of the spectral coefficient
$\{0.05, 0.10, 0.20, 0.30\}$ shows the gain disappears:
all coefficients give 0.795--0.813, within single-seed noise of 0.819 (no spectral).
The original 0.836 was confounded: single seed, and the coefficient was chosen
by looking at the evaluation set.

\bigskip

\section{What the binding constraint is}

Every Conv1D variant plateaus at per-recording BACC $\approx 0.82$,
per-window $\approx 0.77$, regardless of:
\begin{itemize}[noitemsep]
  \item Encoder depth or width (scaling)
  \item 4× more pretraining data (multi-corpus)
  \item Longer pretraining (not tested beyond 20 epochs, but no convergence gap)
  \item SSL objective (VICReg, SIGReg, NaiveJEPA all plateau near the same level)
\end{itemize}

The encoder itself is the ceiling.  The 3.9\,pp gap to LaBraM-Base is not a loss
function problem or a data volume problem — LaBraM uses a learned VQ-codebook
patch tokeniser pretrained on 2\,500 hours of EEG across 12 institutions.
That tokeniser quality, not any SSL objective, is what drives the gap.

\end{document}
