\documentclass[10pt,a4paper]{article}
\usepackage[margin=2cm]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage[expansion=false]{microtype}
\usepackage{enumitem}
\usepackage[T1]{fontenc}

\hypersetup{colorlinks=true,linkcolor=blue!60!black}

\definecolor{warn}{RGB}{192,57,43}
\definecolor{good}{RGB}{39,174,96}

\title{\textbf{Experiment Optimization Plan}\\
  \large Sweeps, Ablations, and Open Decisions for the EB-JEPA EEG Track}
\author{Yani Hammache, Théo Basseras, Titouan Vasnier, Timothy Courtois}
\date{June 2026}

\begin{document}
\maketitle

\begin{abstract}
A ranked catalogue of every sweep, ablation, and methodological fix still open on
this project, organized by expected information gained per GPU-hour rather than by
category. Cross-referenced against \texttt{reports/DECISION\_LOG.md}'s open decisions
(D1--D6) and \texttt{reports/EXPERIMENTS.md}'s ``considered but not run'' list, so
this plan extends existing open questions instead of duplicating them.
\end{abstract}

\section{Statistical rigor (cheapest, currently missing entirely)}
\begin{enumerate}
  \item \textbf{Multi-seed training.} Every encoder/loss config run so far used a
  single seed (\texttt{--single} in the launcher, for speed). The launcher already
  supports \texttt{--sweep <name>} with 3 seeds (1/1000/10000) for free. Re-running the
  top 2--3 configs with 3 seeds would tell us whether ConvNet's lead over LaBraM is
  real or noise.
  \item \textbf{Bootstrap / clustered confidence intervals} on probe metrics. Windows
  from the same recording are not i.i.d.; a naive CI understates uncertainty. Nobody
  has computed this yet.
  \item \textbf{Probe-only seed variance.} Re-fitting the MLP probe with 5 different
  \texttt{random\_state} values costs nothing (no retraining) and separates probe
  variance from encoder variance.
\end{enumerate}

\section{Loss / objective coefficient sweeps}
\begin{itemize}
  \item \textbf{VICReg}: now fixed to the paper's $(\lambda,\mu,\nu)=(25,25,1)$, but
  worth sweeping around it -- e.g.\ $(15,15,1)$, $(40,40,1)$ -- per architecture,
  since the bug's effect was architecture-dependent (catastrophic for EEGPT, mild for
  Conv).
  \item \textbf{SIGReg}: \texttt{num\_slices} $\in \{64,128,256,512\}$,
  \texttt{lmbd} $\in \{1,10,50\}$.
  \item \textbf{Predictive JEPA}: EMA decay \texttt{ema} $\in \{0.99, 0.996, 0.999\}$
  (too fast $\Rightarrow$ target chases online, defeating the point; too slow
  $\Rightarrow$ barely updates); prediction horizon (currently predicts every future
  frame one-by-one from frame 0 -- could instead predict only frame $+1$, or frame
  $+k$ for varying $k$); \texttt{pred\_std\_coeff}/\texttt{pred\_cov\_coeff}.
\end{itemize}

\section{Augmentation ablations (cheap, high diagnostic value)}
\begin{itemize}
  \item \textbf{Leave-one-out}: disable each of the 4 augmentation steps (scale
  jitter, noise, channel-drop, time-mask) individually and retrain, to isolate which
  one actually drives the gain. \texttt{EXPERIMENTS.md} explicitly flags this as never
  separated for the corruption variant either.
  \item \textbf{EEGPT-specific}: disable channel-drop only for EEGPT
  (\textbf{D1} in the decision log -- channel-drop conflicts with EEGPT's
  channel-pooling bottleneck; jobs 75752/75806 test this).
  \item \textbf{Exact-corruption coefficients} (\texttt{aug\_mask\_frac},
  \texttt{aug\_outlier\_frac}, \texttt{aug\_outlier\_scale}): only the spectral-loss
  coefficients (0.05/0.1/0.3) were swept; the corruption-specific knobs themselves
  were not.
\end{itemize}

\section{Architecture hyperparameter sweeps}
\begin{itemize}
  \item Conv: \texttt{hidden} / \texttt{depth} / \texttt{kernel\_size}.
  \item Transformers: \texttt{embed\_dim}, \texttt{tr\_depth}, \texttt{n\_heads},
  \texttt{patch\_len} (1\,s patches were never compared against e.g.\ 0.5\,s or 2\,s).
  \item \textbf{Capacity-matched comparison (D6, open).} Conv ($\sim$0.4M params) is
  currently compared to LaBraM/EEGPT/BIOT at whatever size they happen to be --
  ``architecture wins'' and ``more capacity wins'' are confounded. Scaling every
  encoder to an equal parameter count before comparing is the single biggest
  methodological gap in the architecture comparison.
\end{itemize}

\section{Training schedule}
\begin{itemize}
  \item \textbf{Epoch count}: everything so far is 20 epochs. LaBraM trained
  $\sim$200 epochs on a much larger corpus -- 20 may simply be too few for
  transformers to converge, independent of architecture quality.
  \item \textbf{LR schedule}: currently flat AdamW, no warmup/decay -- a cosine
  schedule could matter more for transformers than for Conv.
  \item \textbf{Batch size}: VICReg's covariance term needs enough samples per batch
  to estimate decorrelation meaningfully; a larger batch could shift the
  variance/covariance balance.
\end{itemize}

\section{Data scale}
\begin{itemize}
  \item \textbf{\texttt{subset\_frac} scaling curve}: 5\% $\to$ 25\% $\to$ 50\%
  $\to$ 100\%, plotted against probe BACC -- reveals whether any encoder is
  data-starved (transformers usually need more data than Conv before showing their
  advantage).
  \item \textbf{Multi-corpus} (TUAB+TUEV+TUSZ+TUEP, tried once on Conv only by
  tvasnier) -- never tried on the transformer encoders, where extra data might
  matter more.
\end{itemize}

\section{Probe / eval protocol}
\begin{itemize}
  \item \textbf{LogReg vs.\ MLP probe (D3, open)}: only MLP has been used so far.
  LogReg is the literature standard, more directly comparable to SimCLR/MoCo-style
  papers, and removes the probe-capacity confound.
  \item \textbf{\texttt{n\_windows} sweep} (8/16/32/all) for per-recording pooling;
  now that per-window scoring exists, compare both protocols at multiple $N$.
  \item \textbf{Re-score every existing checkpoint under both protocols} (per-window
  now implemented but not yet run) for a complete dual table.
\end{itemize}

\section{Cross-component combinations (each tested in isolation so far)}
\begin{itemize}
  \item \textbf{Spectral loss + corrected VICReg (D4, open).} The best single result
  (0.836 per-recording) used the spectral auxiliary loss but with the
  \emph{buggy} VICReg coefficients. Never retrained with the fix -- open question
  whether spectral was adding independent signal or just compensating for the bug.
  \item \textbf{Corruption + SIGReg} (lightly tried) vs.\ \textbf{corruption +
  spectral} (tried, regressed) -- the fuller cross product (augmentation
  $\times$ loss $\times$ anti-collapse mechanism) has not been run.
  \item \textbf{Predictive JEPA + SIGReg} anti-collapse instead of the current
  Hinge-std/covariance pair.
\end{itemize}

\section{Architectural fixes worth isolating}
\begin{itemize}
  \item \textbf{BIOT with phase}: currently FFT-magnitude only (discards phase) --
  adding a complex/phase channel to the token embedding directly tests whether
  phase loss is the bottleneck.
  \item \textbf{EEGPT without the channel-pool bottleneck}: replace the
  channel-summarization attention with simple mean-pool across channels, to test
  whether the hierarchical design itself (not just channel-drop) is the problem.
  \item \textbf{LaBraM-style tokenizer warm-start (D2, open)}: pretrain the patch
  embedding with a quick reconstruction objective before SSL, closer to the real
  LaBraM recipe.
\end{itemize}

\section{Extensions beyond TUAB}
\begin{itemize}
  \item \textbf{TUEV} (6-class, harder, imbalanced) -- the original README's
  suggested follow-up, untouched.
  \item Other datasets already scoped in \texttt{suggestions/datasets/}:
  BCI-IV-2a, HMC, sleep-EDF, HGD -- would test whether any encoder's advantage is
  TUAB-specific or general.
\end{itemize}

\section{Ranked recommendation}
By expected information gained per GPU-hour, in order:
\begin{enumerate}
  \item Multi-seed runs on the existing best 2--3 configs (Section~1).
  \item Capacity-matched architecture comparison (Section~4).
  \item Augmentation leave-one-out (Section~3).
  \item Epoch-count sweep for the transformer encoders (Section~5).
  \item Spectral + fixed-VICReg retrain, i.e.\ D4 (Section~7).
\end{enumerate}
These five directly resolve open questions already on record, rather than opening
new ones -- the highest-leverage use of remaining compute.

\end{document}
