\documentclass[10pt,a4paper]{article}
\usepackage[margin=2cm]{geometry}
\usepackage{booktabs,array,xcolor,graphicx}
\usepackage[T1]{fontenc}
\usepackage[expansion=false]{microtype}
\usepackage[colorlinks=true,linkcolor=blue!60!black]{hyperref}

\definecolor{good}{RGB}{39,174,96}
\definecolor{bad}{RGB}{192,57,43}
\newcommand{\good}[1]{\textcolor{good}{#1}}
\newcommand{\bad}[1]{\textcolor{bad}{#1}}

\title{\textbf{EB-JEPA --- Robustness, Hyperparameter Tuning \& Ablation (V1)}}
\author{Yani Hammache, Théo Basseras, Titouan Vasnier, Timothy Courtois}
\date{June 2026}

\begin{document}
\maketitle

\noindent Five experiments distilled from a 5-stage optimization of our EB-JEPA on TUAB
(v3.0.1, patient-disjoint, 276 eval recordings; frozen encoder $\to$ logistic-regression probe).
Row~1 is the \textbf{tuned model} --- the winner of the campaign (capacity $\to$ augmentation
$\to$ schedule $\to$ loss coefficients $\to$ architecture/data); rows~2--5 each perturb it along
one axis. \emph{per-rec} = per-recording (16-window mean-pool, the clinical metric); \emph{per-win}
= per-window (the literature convention). Single-seed. Companion to
\texttt{reference\_table\_JEPA\_V1} (the full runs table).

\bigskip

\resizebox{\textwidth}{!}{%
\begin{tabular}{@{}llp{6.6cm}ccc@{}}
\toprule
\textbf{\#} & \textbf{Test} & \textbf{Configuration (vs.\ tuned model)} & \textbf{Params}
& \textbf{per-rec BAcc / AUROC} & \textbf{per-win BAcc / AUROC} \\
\midrule
1 & \textbf{Tuned model} & Conv $\cdot$ VICReg + spectral 0.1 + corruption $\cdot$ inv\,=\,1 $\cdot$ no scale-jitter $\cdot$ 20 ep & 1.35M & \textbf{0.824 / 0.896} & 0.755 / 0.839 \\
1b & \textbf{Tuned --- 3-seed} & same recipe, seeds $\{0,1,2\}$ (mean$\pm$std) & 1.35M & 0.812\,$\pm$.021 / 0.898\,$\pm$.002 & 0.756\,$\pm$.001 / 0.836\,$\pm$.005 \\
2 & \textbf{Robustness} & pretrain on \textbf{25\%} of recordings (4$\times$ less data) & 1.35M & \good{0.816} / 0.895 & 0.762 / 0.842 \\
3 & \textbf{HP tuning} & invariance weight \textbf{inv\,=\,25} (VICReg default) instead of 1 & 1.35M & \bad{0.804} / 0.884 & 0.730 / 0.813 \\
4 & \textbf{Ablation} & remove the \textbf{corruption mask} (outliers only) & 1.35M & \bad{0.796} / 0.891 & 0.762 / 0.841 \\
5 & \textbf{Ablation} & \textbf{Transformer} encoder at equal capacity & 1.29M & \bad{0.725} / 0.804 & 0.652 / 0.708 \\
\bottomrule
\end{tabular}}

\bigskip

\noindent\textbf{Legend / fixed setup.} Every row uses the tuned model's recipe except the stated
change: a 0.4M-class 1-D conv scaled to \textbf{1.35M} (\texttt{hidden=96, depth=4}); two-view
VICReg energy + a DDSP multi-scale-spectral auxiliary term (weight 0.1); exact EB-corruption views
$v_1,v_2$ (Gaussian noise $\sigma=0.1$, channel-drop $p=0.2$, 20\% timepoints masked + 20\%
$\pm6\sigma$ outliers); invariance weight 1; 20 epochs; frozen-encoder logreg probe.

\smallskip
\noindent\textbf{What each row shows.}
\begin{itemize}\itemsep1pt
\item[\textbf{1b}] \textbf{Seed robustness (important caveat):} the per-recording BAcc carries
$\approx\pm0.02$ seed noise (only 276 recordings, threshold-sensitive) --- seed 1 dipped to 0.788.
Per-\emph{window} BAcc ($\pm.001$) and AUROC ($\pm.002$) are stable. \emph{Consequence: the per-rec
$\Delta$'s in rows 2--4 ($<3$\,pp) sit within this noise band --- only row 5's $-10$\,pp is
statistically unambiguous. Rank configs on per-window / AUROC, not single-seed per-rec.}
\item[\textbf{2}] \textbf{Robustness to data:} cutting pretraining data 4$\times$ costs only $0.8$\,pp ($0.824\!\to\!0.816$) --- the encoder is \emph{not} data-starved; it saturates early.
\item[\textbf{3}] \textbf{Hyperparameter tuning:} VICReg's vision-default invariance weight (25) is \textbf{2\,pp worse} than our tuned \texttt{inv=1}. EEG SSL prefers \emph{weak} view-invariance (full swept curve: $1>5>10>25$).
\item[\textbf{4}] \textbf{Augmentation ablation:} the corruption \textbf{mask is the single biggest lever} ($-2.8$\,pp when removed); for contrast, scale-jitter and the spectral term are \emph{dead weight} ($\approx0$).
\item[\textbf{5}] \textbf{Architecture ablation:} a from-scratch Transformer at \emph{equal} $\sim$1.3M params is \textbf{10\,pp worse} --- at our scale/epoch budget the conv wins decisively; the SOTA gap is pretraining corpus + schedule, not architecture.
\end{itemize}

\end{document}
