\documentclass[10pt,a4paper]{article}
\usepackage[margin=2cm]{geometry}
\usepackage{booktabs,array,xcolor,graphicx}
\usepackage[T1]{fontenc}
\usepackage[expansion=false]{microtype}
\usepackage[colorlinks=true,linkcolor=blue!60!black]{hyperref}

\title{\textbf{EB-JEPA --- Reference Table (V1)}}
\author{EB-JEPA EEG track}
\date{June 2026}

\begin{document}
\maketitle

\noindent Our EB-JEPA runs on TUAB (v3.0.1, patient-disjoint, 276 eval recordings).
\emph{per-rec} = per-recording (16-window mean-pool, the clinical metric); \emph{per-win} =
per-window (the literature convention). $\pm$ = 3-seed standard deviation; other rows are
single-seed. Companion to \texttt{explanation\_augmented\_corruption.tex} (which details the
augmentation) and \texttt{PROVENANCE\_TABLE.md} (the literature baselines).

\bigskip

\resizebox{\textwidth}{!}{%
\begin{tabular}{@{}llcllcc@{}}
\toprule
\textbf{\#} & \textbf{Encoder} & \textbf{Params} & \textbf{Energy (SSL objective)} & \textbf{Eval}
& \textbf{per-rec BAcc / AUROC} & \textbf{per-win BAcc / AUROC} \\
\midrule
baseline & Conv1D & 0.4M & VICReg --- \emph{base aug, no corruption} & frozen & 0.796 / 0.888 & 0.756 / -- \\
1 & Conv1D & 0.4M & \textbf{VICReg + spectral 0.1} & frozen & 0.836 / 0.887 & 0.765 / -- \\
2 & Conv1D & 0.4M & \textbf{SIGReg} & frozen & 0.825 / 0.913 & \textbf{0.775 / 0.856} \\
3 & Conv1D & 0.4M & \textbf{VICReg} & frozen & \textbf{0.819\,$\pm$.004 / 0.900\,$\pm$.006} & 0.770 / 0.848 \\
4 & Conv1D & 0.4M & VICReg & FT & 0.812\,$\pm$.004 / 0.908\,$\pm$.006 & -- \\
5 & \textbf{Transformer} (multi-corpus) & 3.65M & VICReg & frozen & 0.798 / 0.872 & 0.738 / -- \\
6 & Conv1D & 0.4M & \textbf{VICReg + spectral 0.1 + $\Delta z/\Delta z^2$/ACF/sACF} & frozen & 0.823\,$\pm$.007 / 0.900\,$\pm$.010 & 0.761\,$\pm$.004 / 0.841\,$\pm$.004 \\
\bottomrule
\end{tabular}}

\bigskip

\noindent\textbf{Fixed for every corruption run} (never swept): views $v_1,v_2$ made with Gaussian
noise $\sigma=0.1$, scale-jitter $\pm20\%$, channel-drop $p=0.2$, plus exact EB-corruption ---
20\% timepoints masked + 20\% $\pm6\sigma$ outliers ($\approx$40\%/view). The \textbf{baseline} is
the no-corruption reference (legacy contiguous time-mask).

\medskip

\noindent\textbf{Row 6 --- stylized-facts energy} (3-seed). Row~1's energy plus four
view-consistency blocks between the two views' encoder feature maps, each an MSE between a
path-statistic of $v_1$ and $v_2$, with explicit, fixed a-priori (not eval-swept) coefficients:
increments $\Delta z$ (0.10), realized variance $\mathrm{E}[\Delta z^2]$ (0.05), temporal
autocorrelation over lags 1--8 (0.10), and autocorrelation of the power spectrum / FFT (0.05); the
spectral term stays at its validated 0.10. All $\le0.1$ so VICReg's variance/covariance
anti-collapse terms remain dominant. \emph{Result: squarely on the 0.82 plateau} --- the added
path-statistics regularise the representation but do not lift the 0.4M-conv ceiling (per-window is,
if anything, $\sim$1\,pp lower than row~3).

\end{document}
